---
title: "Assignment Week 1"
output: html_document
---

### Name: Qing Zhou

### ULCN: s2501597

```{r, echo=FALSE}
rm(list=ls(all=T)) #clears your workspace
```


# EXERCISE I. Simulating from the triangular distribution

The *triangular distribution* is the three-parameter family of distributions with probability density looking like this:

```{r, echo= FALSE} 
xa <- -1; xb <- 1; xc <- 0
plot(0, xlim = c(-2,2), ylim = c(0,1.5), type = "n", ylab = "f(x | a,b,c)" , xlab = "x",asp=1)
segments(x0 = xa, x1 = xc, y0 = 0, y1 = 1, lty = 1, lwd = 2)
segments(x0 = xb, x1 = xc, y0 = 0, y1 = 1, lty = 1, lwd = 2)

segments(x0 = xc, y0 = 0, y1 = 1, lty = 2, col = 'red')

```

The three parameters $a$, $b$ and $c$ define the $x$ values at the left, right and top position of the triangle, respectively.
In this assignment we focus on the symmetric case \(c = \frac{1}{2}(a + b)\) and even more in particular, the case \(a = -1, b = +1, c= 0\) which we will call the standard symmetrical triangular distribution.

### (a) Derivation of cdf.

Derive the cumulative distribution function \(F\) of the standard triangular distribution, and the quantile function $Q$.
Add your answers below in this Rmd file.

Hint:

* Make life easy, calculate with \(a = -1, b = 1, c = 0\).

* Rmd allows you to type latex style text, example: \(a = \frac{1}{2}\)

* If you're not familiar with LateX, find a way to put your answer here clearly and concisely.
The pdf of the standard symmetrical triangular distribution can be expressed as
$$f(x|-1,1,0) =  \begin{cases}0, \hfill x<-1\\ x+1, \hfill -1\leq x  <0\\ -x+1,\hfill 0\leq x  < 1 \\ 0,\hfill x\geq 1\end{cases}$$ Its cdf can be found by $$f(x|-1,1,0)  = \frac{dF(x|-1,1,0)}{dx}$$ When $-1\leq x <0$, we have $f(x)=x+1$, and $F(x)=\frac{1}{2}x^2+x+c_1$ where $c_1$ is a constant. With the condition that at $x=0$, $F(x)=\frac{1}{2}$, we can solve for $c_1$ with $c_1=\frac{1}{2}$. Likewise when $0\leq x<1$ we have $f(x)=-x+1$ and $F(x)=-\frac{1}{2}x^2+x+c_2$. It has to  be satisfied that at $x=0,F(x)=\frac{1}{2}$ and $x=1,F(x)=1$. Solve and we find $c_2=\frac{1}{2}$. Hence we obtain $$F(x|-1,1,0)=\begin{cases}0,\hfill x<-1\\\frac{1}{2}x^2+x+\frac{1}{2},\hfill-1\leq x <0\\-\frac{1}{2}x^2+x+\frac{1}{2},\hfill 0 \leq x < 1\\1,\hfill x\geq  1 \end{cases}.$$
```{r, echo= FALSE} 
curve(-1/2*x*abs(x)+x+1/2, -1, 1,xlab="x",ylab="F(x|a,b,c)",lwd=2)
```

When $-1<x\leq 0$, $F(x)=\frac{1}{2}x^2+x+\frac{1}{2}$, and the quantile function can be derived by finding the value for which $\frac{1}{2}x^2+x+\frac{1}{2}=p\implies Q(p)  = \sqrt{2p}-1$. 

When $0<x\leq 1$,$F(x) = -\frac{1}{2}x^2+x+1$, and the quantile function can be found by $-\frac{1}{2}x^2+x+1=p\implies Q(p)  = 1-\sqrt{2-2p}$
```{r, echo= FALSE} 
x1 <- seq(0, 0.5, length.out = 50)
x2 <- seq(0.5,1,length.out  = 50)
y1 <- sqrt(2*x1)-1
y2 <- -sqrt(2-2*x2)+1
plot(x1, y1, xlim=c(0,1),ylim=c(-1,1),type = "l",xlab="p",ylab="Q(p|a,b,c)")
lines(x2, y2, type='l')
```

### (b) Draw sample.

Write R functions to generate a random sample of size \(n\) from the standard triangular distribution;

1. using the trick that the sum of the two independent uniforms has a triangular distribution,

```{r}
start_time <- Sys.time()
x <- runif(100000,-.5,.5)
y  <- runif(100000,-.5,.5)
z <-  x+y
end_time <- Sys.time()

end_time - start_time
hist(z,breaks=100,freq=FALSE)
```

2. using the inverse transform method,

```{r}
start_time <- Sys.time()
x <- runif(100000,0,1)
x1 <- x[x<0.5]
x2 <- x[x>=0.5]
y1 <- sqrt(2*x1)-1  
y2 <- 1-sqrt(2-2*x2) 
y <- append(y1,y2)
end_time <- Sys.time()

end_time - start_time
hist(y,breaks=100,xlim=c(-1,1),freq=FALSE)

```

3. using the acceptance-rejection method (*cf. Rizzo Example 3.7*),

```{r}
start_time <- Sys.time()
u <- runif(100000,0,1)
invcdf.func <- function(u) {
 if (u >= 0 && u < 0.5)
 sqrt(2*u)-1
 else if (u >= 0.5 && u <= 1)
 1 - sqrt(2-2*u)
 }
 x <- unlist(lapply(u, invcdf.func))
 end_time <- Sys.time()

end_time - start_time
hist(x,breaks=100,xlim=c(-1,1),freq=FALSE)
```


4. and last, use the functions `system.time()` and `replicate()` to compare the efficiency of these methods.

See  above. The fastest method is the first one, and the slowest method is the last one.

```{r}

```

# EXERCISE II. Mixtures of Gaussians

$X$, $Y$ and $Z$ are independent, and normally distributed with variance 1; their means are $-1$, $1$ and $0$ respectively. Now define the following two additional random variables:

* $U = (X+Y)/2$
* $V = \begin{cases}
X&\text{if $Z<0$, and}\\
Y&\text{otherwise}.
\end{cases}$

### (a) Plot histograms
Generate a random sample from the distribution of $U$, and plot a historgram using the function `hist`. Do the same for the distribution of $V$.

```{r}
X <- rnorm(1000000,-1,1)
Y <- rnorm(1000000,1,1)
Z <- rnorm(1000000,0,1)
U <- X+Y
V1 <- X[Z<0]
V2 <- Y[Z>=0]
V <- c(V1,V2)
hist(U,breaks=100,freq=FALSE)
hist(V,breaks=100,freq=FALSE)
```

### (b) What about the distributions?
* Write down the densities of $U$ and $V$.

The characteristic functions of X  and Y are $\varphi_X(t) = E(e^{itX})$  and  $\varphi_Y(t) = E(e^{itY})$.  So the characteristic function of $U$ becomes
$$\varphi_{X+Y}(t) = \varphi_X(t)\varphi_Y(t)=\exp\left(-it-t^2/2\right)\exp\left(it-t^2/2\right)=\exp(-t^2)$$
So the distribution of $U$ is a Gaussian and obeys  $U\sim \mathcal{N}(0,2)$.

* Which of these densities is bimodal (has multiple local maxima)?

If we change the variance of $X,Y,Z$ to be 0.5, then the result is more obvious

```{r}
X <- rnorm(1000000,-1,.5)
Y <- rnorm(1000000,1,.5)
Z <- rnorm(1000000,0,.5)
U <- X+Y
V1 <- X[Z<0]
V2 <- Y[Z>=0]
V <- c(V1,V2)
hist(U,breaks=100,freq=FALSE)
hist(V,breaks=100,freq=FALSE)
```
So the  distribution of $V$ is bimodal.

* Which of these densities describes a mixture of the distributions of $X$ and $Y$?

The distribution of $V$ can be described by a mixture of $X$ and $Y$.

* What distribution is described by the other density?

The distribution of $U$ is a normal distribution, and the distribution of $V$ is a mixture of two normal distributions, and is a bimodal distribution. So the bimodal distribution can be described by two normal distributions.
